BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets

### 2024-08-16 manuscript/manuscript_APPS.rmd
writing about other software

--- 
title: 'BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets'
author:  |
    | Reed Clark Benkendorf$^1$^[Author for Correspondence: rbenkendorf@chicagobotanic.org], Jeremie B. Fant$^1$$^,$$^2$
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
abstract: |
  **Premise:** Depositing specimens to herbaria is a time consuming task. Many institutions have reduced the amount of funding for herbaria, and universities have reduced the amount of education dedicated to curatorial tasks and specimen deposition. Despite this, the continual generation of herbaria specimens are essential for current and future research in evolution and ecology. In order to faciliate the continued growth of herbaria BarnebyLives was developed as tool to supplement collection notes, perform geographic and, taxonomic informatic processes, enact spell checks, produce labels, and submit digital data.    
  **Methods and Results:** BarnebyLives uses geospatial data from the U.S. Census Bureau to provide political jurisdiction information, and data from other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions. It uses inhouse spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclatural innovations. Optionally the package writes driving directions to sites using Google Maps. The package outputs data in a tabular format for review by the user to accept or confirm changes, before dynamically rendering labels.       
  **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users the most current taxonomic opinions, generates driving directions to sites, and produces aesthetically appealing labels and shipping manifests in a matter of minutes.     
keywords: |
  herbariaum, natural history museum, collections, geospatial, automation, R, software 
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/american-journal-of-botany.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
- \usepackage[width=\textwidth]{caption}
- \usepackage{wrapfig}
- \usepackage[export]{adjustbox}
---

@@

\begin{wrapfigure}{l}{0.35\textwidth}
  \centering
    \includegraphics[max size={\textwidth}{\textheight}]{../graphics/plots/flowchart-trim.png}
  \caption{Recommended workflow.}
\end{wrapfigure}

@@

@@

```{r install package}
## devtools::install_github('sagesteppe/BarnebyLives')
```

@@

time_split_binomials <- system.time({ # split up names into their components
  data <- split_binomial(data, 'Scientific_name')
})

@@

BarnebyLives took roughly `r minutemen(sum( time_trials[time_trials$Group=='Local', 'elapsed']))` minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`s) to run all local steps, and roughly `r minutemen(time_trials[time_trials$Function=='powo_searcher', 'elapsed'])` minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`s) to search Plants of the World Online, and `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']`s to search Google Maps and write directions to sites. 
Most of the local run time is attributable to the spatial (spatial: `r sum(tt_local[tt_local$Module=='Geospatial', 'elapsed'])`s), and taxonomic operations (`r sum(tt_local[tt_local$Module=='Taxonomic', 'elapsed'])`s), style: `r sum(tt_local[tt_local$Module=='Style', 'elapsed'])`s. 
The spell check operation of the scientific name accounted for nearly all of the time (`r sum(tt_local[tt_local$Function=='spell_check', 'elapsed'])`s) spent performing local taxonomic operations.
The generation of labels consumed around `r minutemen(time_label_gen)` minutes (`r time_label_gen`s) for the rendering, `r time_4perpage`s to combine individual labels four per single sheet of landscape orientated paper, and `r time_final`s to combine the `r length(list.files('labels/processed'))` sheets to a single Portable Document Format (PDF).  
The total computer run time for processing these `r nrow(data)` specimens was `r minutemen(sum(time_trials$elapsed))` minutes.

@@

## Results

@@

@@

# AUTHOR CONTRIBUTIONS
The project was conceptualized by R.C.B. The program was written by R.C.B.
Data collection and analysis were performed by R.C.B. 
R.C.B. wrote the manuscript with input from all other authors. 
All authors approved the final version of the manuscript.

>>>>

--- 
title: 'BarnebyLives: an R package to create herbarium specimen labels and clean spreadsheets'
author:  |
    | Reed Clark Benkendorf$^1$^[Author for Correspondence: rbenkendorf@chicagobotanic.org], Jeremie B. Fant$^1$$^,$$^2$
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
abstract: |
  **Premise:** Depositing specimens to herbaria is a time consuming task. Many institutions have reduced the amount of funding for herbaria, and universities have reduced the amount of education dedicated to curatorial tasks and specimen deposition. Despite this, the continual generation of herbaria specimens are essential for current and future research in evolution and ecology. In order to faciliate the continued growth of herbaria BarnebyLives was developed as tool to supplement collection notes, perform geographic and, taxonomic informatic processes, enact spell checks, produce labels, and submit digital data.    
  **Methods and Results:** BarnebyLives uses geospatial data from the U.S. Census Bureau to provide political jurisdiction information, and data from other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions. It uses inhouse spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclatural innovations. Optionally the package writes driving directions to sites using Google Maps. The package outputs data in a tabular format for review by the user to accept or confirm changes, before dynamically rendering labels.       
  **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users the most current taxonomic opinions, generates driving directions to sites, and produces aesthetically appealing labels and shipping manifests in a matter of minutes.     
keywords: |
  herbariaum, natural history museum, collections, geospatial, automation, R, software 
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/american-journal-of-botany.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
- \usepackage[width=\textwidth]{caption}
- \usepackage{wrapfig}
- \usepackage[export]{adjustbox}
- \usepackage{rotating}
---

@@

@@

The label generation functionality is provided explicitly by two other programs PLabel, and Symbiota, as well as commonly by the Microsoft Word tool Mail Merge (@perkins2020Plabel, @gries2014symbiota). 
The office suite costs money, and in our experience is notoriously finicky, further it's functionality ends with label creation; while we expect this tools exists in open source alternatives such as Libreoffice and OpenOffice, we expect them to have similar issues with use. 
PLabel has greatly enhanced functionality relative to a Mail Merge, allowing users to specify the layout and formatting of label components using an intuitive and local graphical user interface functionality, unfortunately it does not include data cleaning functionality; while some sources indicate it can only be used on Microsoft we expect it can be accessed on Linux and Mac using Windows emulators like Wine. 
The popular Symbiota biodiversity data management software not only provides label generation capabilities but also provides data cleaning functionality, in an attractive GUI web portal allowing for live management of collections. 
However, not all herbaria use Symbiota and many have original database systems which they maintain. 
Furthermore, many collectors like to process their own labels, especially as they are likely to be sending sets to different institutions. 
Accordingly, Symbiota's functionality should exist in an ecosystem with alternative systems.

@@

\includegraphics[height=\textheight, angle=270]{../graphics/plots/workflow.png}

```{r install package}
## devtools::install_github('sagesteppe/BarnebyLives', force = TRUE)
```

@@

time_split_binomials <- system.time({ # split up names into their components
  data1 <- split_scientificName(data, 'Scientific_name')
})

@@

BarnebyLives took roughly `r minutemen(sum( time_trials[time_trials$Group=='Local', 'elapsed']))` minutes (`r sum( time_trials[time_trials$Group=='Local', 'elapsed'])`s) to run all local steps, and roughly `r minutemen(time_trials[time_trials$Function=='powo_searcher', 'elapsed'])` minutes (`r time_trials[time_trials$Function=='powo_searcher', 'elapsed']`s) to search Plants of the World Online, and `r time_trials[time_trials$Function=='directions_grabber', 'elapsed']`s to search Google Maps and write directions to sites. 
Most of the local run time is attributable to the spatial (`r sum(tt_local[tt_local$Module=='Geospatial', 'elapsed'])`s), and taxonomic operations (`r sum(tt_local[tt_local$Module=='Taxonomic', 'elapsed'])`s), style: `r sum(tt_local[tt_local$Module=='Style', 'elapsed'])`s. 
The spell check operation of the scientific name accounted for nearly all of the time (`r sum(tt_local[tt_local$Function=='spell_check', 'elapsed'])`s) spent performing local taxonomic operations.
The generation of labels consumed around `r minutemen(time_label_gen)` minutes (`r time_label_gen`s) for the rendering, `r time_4perpage`s to combine individual labels four per single sheet of landscape orientated paper, and `r time_final`s to combine the `r length(list.files('labels/processed'))` sheets to a single Portable Document Format (PDF).  
The total computer run time for processing these `r nrow(data)` specimens was `r minutemen(sum(time_trials$elapsed))` minutes.

@@

## RESULTS

@@

# DISCUSSION

While numerous tools have been developed for cleaning of existing herbarium and museum records, few help with ensuring that the entered data are accurate (@patten2024geographic),

Accessioning often times relies on the use of the Office Suite of programs, and may utilize other costly softwares, such as ArcPro, or Adobe Acrobat. 
While BarnebyLives does not have it's own graphic user interface, the functionality of commonly used Interactive Development Environments (IDE's), such as Rstudio and VisualStudio (VS) Code, now offer functionality to readily view and filter data sets using familiar spreadsheet formats

LaTeX offers well documented and detailed functionality for customizing labels for individuals and institutions... 
While the program

@@

# AUTHOR CONTRIBUTIONS
The project was conceptualized by R.C.B. The program was written by R.C.B.
Data collection and analysis were performed by R.C.B. 
R.C.B. & J.B.F wrote the manuscript, and both authors approved the final version of the manuscript.
